Current Affairs: Utilizing Oxford Nanopore Sequencing Data to Detect Non-Canonical DNA Structures

### 2024-01-26 jan30meeting_pres1/may30ccgmeeting.tex
updates

	% For my Background slides, I will talk about important things such as WHat LCRs are, their evolution, stuff like that
	% SLIDE 3 - WHAT ARE LCRs%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	% The graphic for this slide is an image of some output from segA with the parameters in the brackets
	\section{Background}
	\begin{frame}{A Brief History Lesson}
		\begin{itemize}
			\item (1922) American geneticist Calvin Bridges was crossing fruit flies (\textit{Drosophila melanogaster}) and observed certain genes were lethal only in combination \newline
			\item (1946) Years later, a colleague of Bridges observed the same phenomena in a different species of fruit fly, \textit{Drosophila pseudoobscura}, and coined the term `synthetic lethality' \newline
		\end{itemize}

		\centering
		\huge \textbf{Synthetic = Combining}

@@

	%Also called synthetic sick because it can lead to reduced cell viability as opposed to death
	\begin{frame}{The Combined Inactivation of Two Genes Can Lead to Synthetic Lethal Interactions}
		\label{link1}

		\begin{itemize}
			\item Synthetic lethal interactions describe the relationship between two genes whose coupled inactivation, but not their individual inactivation, causes cell death or reduces cell viability \newline

			\includegraphics[width=11cm, height=2.5cm]{slbasic.png}

			\item Inactivation: Preventing or disabling normal function of a gene (e.g mutation)
			\footnotetext[2]{\tiny\cite{lee2018harnessing}}
		\end{itemize}

@@

	%SLIDE 5 - LCR's PRESENT IN UNIQUE WAYS
	\begin{frame}{Researchers Harness Synthetic Lethal Interactions in a Variety of Biological Applications}

		\begin{block}{\textit{To gain insight into gene function}}
			 \begin{itemize}
			 \item In \textit{Saccharomyces cerevisiae} (baker's yeast), the mapping of synthetic lethal interactions has reached a genome-wide scale and these networks of interaction are a rich source for the functional annotation of genes
			 %The elucidation of similar comprehensive synthetic lethal interaction networks in human cells would be very attractive, particularly as a large number of genes in the human genome have still not been defined a function
			 \end{itemize}
		\end{block} \pause

		\begin{block}{\textit{To uncover new opportunities for precision oncology}}
			\begin{itemize}
				\item An emerging strategy is to identify potential synthetic lethal interactions between tumor suppressor genes and other genes that can be simultaneously disrupted leading to cancer cell death (e.g. Olaparib was the first drug to work via a synthetic lethal mechanism)
				%Rather than attempting to restore function of mutated tumor suppressor genes (TSGs) which proves extremely difficult to do,
				%Talk about BRCA1 and PARP inhibitors for olaparib lol
			\end{itemize}
		\end{block}

@@

	%Slide 6
	% Poly(ADP-ribose) polymerases (PARPs) repair DNA SSBs through the BER pathway. PARP inhibitors, such as olaparib, prevent repair by trapping the inactivated PARP onto the SSB, resulting in the generation of DNA DSBs during the replication process. In tumors with a homologous recombination deficiency (HRD), such as a BRCA1/2 mutation, the low-fidelity repair mechanism of NHEJ leads to increasing genetic instability and ultimately death of the tumor cell.
	\begin{frame}{Four FDA Approved Anti-Cancer Drugs Work via a Synthetic Lethal Mechanism}
		\begin{itemize}
			\item Olaparib, Niraparib, Rucaparib, Talazoparib are Poly [ADP-ribose] polymerase 1/2 (PARP1/2) inhibitors
		\end{itemize}
		\centering
		\includegraphics[width=8.5cm, height=6cm]{olaparib2.png}

@@

	%SLIDE 6 
	\begin{frame}{Synthetic Lethality can also Help Explain Factors Related to the Tissue-Specificity of Cancer}

	\textbf{\color{blue}Tissue Specificity:} Cancers which manifest in different human tissues have different molecular, phenotypic, and epidemiological characteristics. Some aspects of this phenomena are; \newline

	\begin{enumerate}
		\item Variation in lifetime cancer risk
		\item Variation in cancer onset age
		\item Variation in the genes driving the cancer across tissue types \newline
	\end{enumerate}
		The literature has only reported the \textbf{\color{blue}number of lifetime stem cell divisions} and the \textbf{\color{blue}abnormal levels of CpG island DNA methylation} to be factors able to explain the variance in lifetime cancer risk across tissues (aside from well known global and cancer-specific factors)
	\end{frame}

	\begin{frame}{Quantifying cSL Load in Normal Human Tissue}
			\textbf{\color{blue}cSL Load:} Quantitatively captures the level of cancer synthetic lethal (cSL) gene pair co-inactivation based on gene expression data from normal human tissues
			\begin{center}
				\includegraphics[width=10cm, height=5.7cm]{csl.png}
			\end{center}
		\footnotetext[5]{\tiny Adapted From  \cite{cheng2021synthetic}}
	\end{frame}

@@

	\begin{frame}{Tissue cSL Load in Normal Tissues is Negatively Correlated with Lifetime Cancer Risk}
		\textbf{\color{blue}Tissue cSL Load:} Median cSL load value across all samples of each tissue type

@@

		\begin{frame}{Synthetic Lethality Increases Predictive Power of Models Associated with Tissue Specificity}
		\begin{columns}
			\begin{column}{0.5\textwidth}
					\includegraphics[width=7cm, height=3.5cm]{tcl2.png}
					\centering
					Stem Cell Divisions
			\end{column}
			\begin{column}{0.5\textwidth}
				\includegraphics[width=6cm, height=3cm]{tcl3.png}
				\centering
				\vspace{-0.5cm}Methylation
			\end{column}
		\end{columns}	
		\centering	
	\footnotetext[5]{\tiny\cite{cheng2021synthetic}}
	\end{frame}

>>>>

	% For my Background slides, I will talk about important things such as WHat LCRs are, their evolution, stuff like that
	% SLIDE 3 - WHAT ARE LCRs%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	% The graphic for this slide is an image of some output from segA with the parameters in the brackets
	\section{Background}
	\begin{frame}{The Structure of DNA: A Brief History Lesson}
		\begin{itemize}
			\item In 1953, Watson, Crick, Wilkins, Franklin, and Gosling were the first to describe the structure of DNA \newline
			\item They discovered the \textcolor{blue}{right-handed double helix} (canonical B-form DNA), the most common form found in cells \newline
		\end{itemize}

		\centering
		\includegraphics[width=11cm, height=3.5cm]{rosy.png}

@@

	%Also called synthetic sick because it can lead to reduced cell viability as opposed to death
	\begin{frame}{DNA Can Adopt Alternative Structures}
		\label{link1}

		\begin{itemize}
			\item Now, more than 15 types of DNA structure that differ from the canonical B-form have been reported (non-canonical or non-B form DNA) \newline
			\item Through sequencing of the human genome, we now know over half the genome is composed of repetitive elements - these were initially thought to be `junk DNA' \newline

			\begin{center}
				\includegraphics[width=9cm, height=1.2cm]{repeats.png}
			\end{center}

			\item A crucial feature of some repetitive sequences is their ability to fold into non-canonical DNA structures (non-B DNA)
			\footnotetext[2]{\tiny\cite{lee2018harnessing}}
		\end{itemize}

@@

	%SLIDE 5 - LCR's PRESENT IN UNIQUE WAYS
	\begin{frame}{Types of Non-canonical DNA structures}

		\centering
		\includegraphics[width=9cm, height=6.5cm]{nonb_types.png}

@@

	%Slide 6
	% Poly(ADP-ribose) polymerases (PARPs) repair DNA SSBs through the BER pathway. PARP inhibitors, such as olaparib, prevent repair by trapping the inactivated PARP onto the SSB, resulting in the generation of DNA DSBs during the replication process. In tumors with a homologous recombination deficiency (HRD), such as a BRCA1/2 mutation, the low-fidelity repair mechanism of NHEJ leads to increasing genetic instability and ultimately death of the tumor cell.
	\begin{frame}{Non-Canonical DNA Structures are Involved in Biological Processes}

		Non-B DNA structures have been shown to co-localize with \textcolor{blue}{functional genomic loci} (promoters, enhancers, etc) and \textcolor{blue}{genetic instability hotspots} \newline

		This suggests a role for non-B DNA in vital cellular events such as;
		\begin{itemize}
			\item Regulation of transcription
			\item Regulation of DNA replication and recombination
			\item Regulating genome integrity
		\end{itemize}

		\centering
		%\includegraphics[width=8.5cm, height=6cm]{olaparib2.png}

@@

	%SLIDE 6 
	\begin{frame}{Diseases Associated with Non-Canonical DNA structures}

	\textbf{\color{blue}Repeat Expansion Diseases:} Expansions of non-B DNA structure-forming repeats have been implicated in many neurodegenerative and neuromuscular diseases. \newline

	\textbf{\color{blue}Genetic Instability Diseases:} Non-canonical DNA structures are associated with increased mutability (point mutations, deletions, insertions and chromosomal translocations)

	\begin{itemize}
		\item Enriched at chromosomal breakpoints in translocation-related cancers such as lymphomas and leukaemias.
		\item Can be recognized by DNA repair proteins, triggering error-generating repair processes
		\item G-quadruplexes are present within most human oncogenic promoters and at telomeres - a current theraputic target to downregulate transcription or block telomere elongation in cancer cells.
	\end{itemize}

	\begin{frame}{How are Non-B Structures Detected in the Genome?}
			\begin{center}
				\vspace{-1cm}
				\includegraphics[width=12cm, height=8cm]{Predicting_nonb_structures.png}
				\vspace{-1.5cm}
			\end{center}
		These approaches are based primarily on \textcolor{blue}{DNA sequence motifs}, which are \textcolor{blue}{necessary}, but \textcolor{blue}{insufficient} for formation and are not available for all non-B DNA structures

@@

	\begin{frame}{A Promising New Approach}
		\textbf{\color{blue}Tissue cSL Load:} Median cSL load value across all samples of each tissue type

@@

	\begin{frame}{Oxford Nanopore Sequencing Technology}
		\centering
		\begin{figure}[!htb]
			\centering
			\begin{minipage}{.3\textwidth}
				\centering
				\includegraphics[width=3.5cm, height=6cm]{ont_machine.jpg}
				\caption{ONT Sequencer}
				\label{fig:prob1_6_2}
			\end{minipage}%
			\begin{minipage}{0.7\textwidth}
				\centering
				\movie[width=8cm, height=4.5cm]{placeholder box}{ont.mp4}
				\caption{Inside the Nanopore}
				\label{fig:prob1_6_1}
			\end{minipage}
		\end{figure}

	\footnotetext[5]{\tiny\cite{cheng2021synthetic}}
	\end{frame}
